---
title: "Lab 2: Why Start With Indicators?"
author: "Matt Harris & Michael Fichman"
date: '2023-09-18'
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lab2 Content?
### Introduction and Setting
#### Introduction

from Steif (2021), [Section
1.1](https://urbanspatial.github.io/PublicPolicyAnalytics/TOD.html#why-start-with-indicators):
This is an indicator - a stylized fact that gives simple insight into a
complicated phenomena. Mastering indicators is critical for conveying
nuanced context to non-technical audiences. Here are four suggestions on
what makes a good indicator:

-   A *relatable* indicator is typically motivated by a pressing policy
    concern. "How is it possible that passenger rail in New York City
    has such widespread delays, service suspensions, and rider
    discontent?" A great indicator solicits interest from an audience.

-   A *simple* indicator may be used as an exploratory tool in place of
    more complex statistics. Simplicity helps the audience understand
    the indicator's significance and keeps them engaged in the analysis.

-   A *relative* indicator draws a contrast. "How can New York City
    passenger rail, with the most trips, still loose more money than
    each of the next ten largest cities?" Contextualizing an indicator
    with a relevant comparison makes for greater impact.

-   A *good* indicator typically generates more questions than answers.
    Thus, a good indicator fits into a broader narrative which helps
    motivate a more robust research agenda and ultimately, more applied
    analytics.

In this lab we will walk through:
  1.Gather Data: Use {tidycensus} to gather ACS data for two different years
  2.Visualize Data: Visualize data and maps with {ggplot2}
  3.Geospatial Operations: Use the {sf} package for geospatial operations such as **union**,
    **intersection** and **buffer**
  4.Data Analyze: Use maps and tables to analyze change over time *given* the products
    of the geospatial operations

#### Load packages and functions

```{r setup_packages, warning = FALSE, message = FALSE}
# Load Libraries

library(tidyverse)
library(tidycensus)
library(sf)
library(kableExtra)

options(scipen=999)
options(tigris_class = "sf")

source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")

palette5 <- c("#f0f9e8","#bae4bc","#7bccc4","#43a2ca","#0868ac")
```

```{r load_key, warning = FALSE, eval = FALSE}
census_api_key("c2c1421b5c48d7f3c3c13df40f921a3dd1a940a1", overwrite = TRUE, install = TRUE)
```


### 1- Acs Data Preparation
#### 1-1 Get 2016 ACS data

Methods 1: get "long data" first then transfer it into "wide" data

Step1: Get "long" data
```{r results='hide'}
tracts16 <-  
  get_acs(geography = "tract",
          variables = c("B25026_001E","B02001_002E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E", "B25058_001E",
                        "B06012_002E"), 
          year=2016, state=42,
          county=101, geometry=TRUE) %>% 
  st_transform('ESRI:102728')
```

Step2: Spread the data with `spread()`
```{r}
# Let's "spread" the data into wide form

tracts16 <- 
  tracts16 %>%
  dplyr::select( -NAME, -moe) %>%
  spread(key = variable, value = estimate) %>%
  rename(TotalPop = B25026_001, 
         Whites = B02001_002,
         FemaleBachelors = B15001_050, 
         MaleBachelors = B15001_009,
         MedHHInc = B19013_001, 
         MedRent = B25058_001,
         TotalPoverty = B06012_002)
```

Step3: Create new rate variables using mutate
```{r}
tracts16 <- 
  tracts16 %>%
  mutate(pctWhite = ifelse(TotalPop > 0, Whites / TotalPop, 0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop), 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         year = "2016") %>%
  dplyr::select(-Whites,-FemaleBachelors,-MaleBachelors,-TotalPoverty)
```


#### 1-2 Get 2020 ACS data

Method 2: Get "wide" data in the first place and put the entire process into one continuous sequence

Special functions: `get_acs()`'s agument: `output="wide"`

```{r results='hide'}
tracts20 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E"), 
          year=2020, state=42, county=101, 
          geometry=TRUE, output="wide") %>%
  st_transform('ESRI:102728') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         year = "2020") %>%
  dplyr::select(-Whites, -FemaleBachelors, -MaleBachelors, -TotalPoverty) 
```

#### 1-3 Combine 2016 and 2020 ACS data

Function used here: `rbind()` | The "r" in `rbind()` stands for "row"

```{r}

allTracts <- rbind(tracts16,tracts20)
```

#### Other: Wide data vs long data (and `spread`vs `gather`)

Take a moment to think about the *shape* of data. The gif below is a
visualization of **wide** vs. **long** data. The gif shows the same data
transformed into two different shapes. You will do this often in when
working with data. **Wide** data is a more common representation and is
good for all forms of data analysis, including spatial data. In this
data the `id` is not repeated in the table. On the other hand, **long**
data is takes all the wide-data columns and makes them rows by repeating
the `id` for each piece of data. The **long** format is often good for
plotting and analytics over groups within the data; not as great with
spatial data.

![wide vs long
data](https://www.garrickadenbuie.com/project/tidyexplain/images/tidyr-spread-gather.gif)

By default, `get_acs()` returns **long** data where the `GEOID` is
repeated for each of the acs `variables` requested. The `output`
argument in `get_acs()` can be used to return data in a **wide** format.
Or we can do that ourselves with the `spread()` and `gather()`
functions.

Let's examine each variable and the elements of an sf object

```{r}
glimpse(tracts16)
```

We create a new data frame consisting only of population

```{r}

totalPop16 <-
  tracts16 %>%
  filter(variable == "B25026_001")
```

Ways to examine the data

```{r}
nrow(totalPop16)

names(totalPop16)

head(totalPop16)

glimpse(totalPop16)
```

#### Other: Using ggplot to map census data with {sf} | ??? don't understand

Each plot adds more and more nuance and information

Examine each to see what we've added each time

Consult the textbook to understand the symbology schemes.

(Note: the `qBr()` function used here is an update to the one used
version used in the text book so the results may appear different.)

```{r}
A <- 
  ggplot() +
  geom_sf(data = totalPop16, aes(fill = estimate)) +
  theme(
    plot.title = element_text(size=22)
    )

B <- 
  ggplot() +
  geom_sf(data = totalPop16, aes(fill = q5(estimate))) +
  theme(plot.title = element_text(size=22)) 

C <-
  ggplot() +
  geom_sf(data = totalPop16, aes(fill = q5(estimate))) +
  scale_fill_manual(values = palette5,
                    labels = qBr(totalPop16, "estimate"),
                    name = "Total\nPopluation\n(Quintile Breaks)") +
  theme(plot.title = element_text(size=22))

D <- 
  ggplot() +
  geom_sf(data = totalPop16, aes(fill = q5(estimate))) +
  scale_fill_manual(values = palette5,
                    labels = qBr(totalPop16, "estimate"),
                    name = "Popluation\n(Quintile Breaks)") +
  labs(title = "Total Population", subtitle = "Philadelphia; 2016") +
  mapTheme() + 
  theme(plot.title = element_text(size=22))
```

### 2- Transit Open Data Preparation
#### 2-1 Get Dataframe of two subway lines of Philly
5 Steps to get the whole dataframe: 
  1. get data with `st_read`
  2. change original `line` into the subway name itself
  3. select only two columns: `Station` and `Line`
  4. rebind two dataframe into one with `rebind`
  5. transform the dataframe's crs into the same as `tracts16` with `st_transform`

```{r}

el <- st_read("https://opendata.arcgis.com/datasets/8c6e2575c8ad46eb887e6bb35825e1a6_0.geojson")
Broad_St <- st_read("https://opendata.arcgis.com/datasets/2e9037fd5bef406488ffe5bb67d21312_0.geojson")

septaStops <- 
  rbind(
     el %>% 
      mutate(Line = "El") %>%
      dplyr::select(Station, Line),
     Broad_St %>%
      mutate(Line ="Broad_St") %>%
      dplyr::select(Station, Line)) %>%
  st_transform(st_crs(tracts16))  
```

Visualize it: 

```{r}

ggplot() + 
  geom_sf(data=st_union(tracts16)) +
  geom_sf(data=septaStops, 
          aes(colour = Line), 
          show.legend = "point", size= 2) +
  scale_colour_manual(values = c("orange","blue")) +
  labs(title="Septa Stops", 
       subtitle="Philadelphia, PA", 
       caption="Figure 2.5") +
  mapTheme()
```

#### 2-2 Create buffers of both each stop and union stops

```{r}
# get buffers of each stop
stopBuffer <- st_buffer(septaStops, 2640)

# get one union buffer of whole stops
stopUnion <- st_union(st_buffer(septaStops, 2640))

# create a whole dataframe combines two dataframes above
septaBuffers <- 
  rbind(
     stopBuffer %>%
      mutate(Legend = "Buffer") %>%
      dplyr::select(Legend),
     stopUnion %>%
      st_sf() %>%
      mutate(Legend = "Unioned Buffer"))
```

Visualize both two kinds of data via `facet_wrap()`

PS: Material to know more about `facet_wrap()`: `https://www.juiceanalytics.com/writing/better-know-visualization-small-multiples`

```{r}
ggplot() +
  geom_sf(data=septaBuffers) +
  geom_sf(data=septaStops, show.legend = "point") +
  facet_wrap(~Legend) + 
  labs(caption = "Figure 2.6") +
  mapTheme()
```

#### 2-3 Buffer Choose

In this experience, we only choose the union buffer
```{r}
buffer <- filter(septaBuffers, Legend=="Unioned Buffer")
```

### 3- Spatial Operation about intersection of  Acs data and Transit Data

Mainly use the `sf` Package

Main purpose: Select the tracts from `tracts16` that intersect the buffer of transit stop in `buffer` 

We will try three different ways:
  1. Clipping: The intersection returns only the areas where both `buffer` and `tracts16` polygons overlap.
  2. Touch or intersect: The intersection polygons are returns the full extend of any polygon in `tracts16` that touch (or intersect) the `buffer`.
  3. Centroids: The intersection of polygons to centroids returns the full extent of any polygon in `tracts16` that has its center (called a centroid) within the `buffer`


#### 3-1 Clipping with `st_intersection()` on polygons

```{r}
clip <- 
  st_intersection(buffer, tracts16) %>%
  dplyr::select(TotalPop) %>%
  mutate(Selection_Type = "Clip")
```

#### 3-2 Touch or Intersection with `st_intersects()` on polygons

Simply, this type of spatial selection returns any polygon that overlaps
or touches the buffer. We try to keep our code examples concise in this
course, but here we show three different ways to achieve the same
outcome.

1.  `selection1` uses `tracts16[buffer,]` uses the least code, but is
    often confusing unless you know the {sf} package well. In this
    usage, the intersecting rows of `tracts16` are selected for by
    indexing on `buffer`. This method is very fast to compute, but hard
    to understand at first.

2.  `selection2` is a longhand way to do the same. This uses
    `tracts16[st_intersects(tracts16, buffer) %>% lengths > 0, ]` to
    spell out all of the steps used in the first method. This can be
    read as: Select all rows of the rows from `tracts16` where
    `st_intersects()` returns a length of greater than zero. Important
    note that this uses `st_intersects()` and not `st_intersection()` as
    in the previous section. This approach is just as fast as approach
    #1, but a little easier to read.

3.  `selection3` takes a different approach and uses `st_join()` with the join type of "st_intersects". In this approach we need to follow the join with `filter(!is.na(Legend))` to remove the rows of `tracts16` where the polygons did not join. These are the polygons that did not intersect. This approach is the most readable and easiest to understand, but can be slower on large amounts of data.

```{r spatialSelection}

# Do a spatial selection to see which tracts touch the buffer

# approach #1: sub-setting a spatial object with a spatial object using the '[' brackets.
selection1 <- 
  tracts16[buffer,] %>%
  dplyr::select(TotalPop) %>%
  mutate(Selection_Type = "Spatial Selection")

# approach #2: using `st_intersects` as a verbose way to do approach #1
selection2 <- tracts16[st_intersects(tracts16, buffer) %>% lengths > 0, ] %>%
  dplyr::select(TotalPop) %>%
  mutate(Selection_Type = "Spatial Selection")

# approach #3: use `st_join` to do a spatial join and remove the non-intersecting polygons
selection3 <- tracts16 %>% 
  st_join(buffer, join = st_intersects) %>% 
  filter(!is.na(Legend)) %>% 
  dplyr::select(TotalPop) %>%
  mutate(Selection_Type = "Spatial Intersects")
```

#### 3-3 Centroids of Polygon  with `st_centroid()`
Five steps:
  1. convert tract polygons to centroids: and to the spatial intersect to buffers using the `selection1` approach from above. The resulting geometry type is now a point and not a polygon
  2. drop the geometry: so that is becomes a regular non-spatial data frame
  3. join the original `tracts16` data frame: using the `GEOID` as the join field. Note that only the `GEOID` column is selected, however the polygon spatial features are also joined.
  4. turn back to an {sf} spatial object: `st_sf()` is a function to turn it back to an {sf} spatial object
  5. Get the outcome: finally the `TotalPop` column is selected and a new field of `Selection_Type` is added
  
```{r}
selectCentroids <-
  st_centroid(tracts16)[buffer,] %>%
  st_drop_geometry() %>%
  left_join(., dplyr::select(tracts16, GEOID), by = "GEOID") %>%
  st_sf() %>%
  dplyr::select(TotalPop) %>%
  mutate(Selection_Type = "Select by Centroids")
```

#### 3-4 Plotting the results of each method

```{r}
intersections <- rbind(clip, selection1, selectCentroids)

ggplot() +
  geom_sf(data=intersections, aes(fill = TotalPop)) +
  geom_sf(data=septaStops, show.legend = "point") +
  scale_fill_viridis_c() +
  facet_wrap(~Selection_Type) + 
  mapTheme()
```

## Bonus Exercise
Create a small multiple map of the three types of operations? (NOT GRADED)
### Indicator Maps

We do our centroid joins as above, and then do a "disjoin" to get the
ones that *don't* join, and add them all together. Do this operation and
then examine it. What represents the joins/doesn't join dichotomy? Note
that this contains a correct 2016-2020 inflation calculation

```{r}
allTracts.group <- 
  rbind(
    st_centroid(allTracts)[buffer,] %>%
      st_drop_geometry() %>%
      left_join(allTracts) %>%
      st_sf() %>%
      mutate(TOD = "TOD"),
    st_centroid(allTracts)[buffer, op = st_disjoint] %>%
      st_drop_geometry() %>%
      left_join(allTracts) %>%
      st_sf() %>%
      mutate(TOD = "Non-TOD")) %>%
  mutate(MedRent.inf = ifelse(year == "2009", MedRent * 1.14, MedRent)) 

```

Can you try to create the maps seen in the text? The solutions are
contained in "map_exercise.R"

### TOD Indicator Tables

```{r}
allTracts.Summary <- 
  st_drop_geometry(allTracts.group) %>%
  group_by(year, TOD) %>%
  summarize(Rent = mean(MedRent, na.rm = T),
            Population = mean(TotalPop, na.rm = T),
            Percent_White = mean(pctWhite, na.rm = T),
            Percent_Bach = mean(pctBachelors, na.rm = T),
            Percent_Poverty = mean(pctPoverty, na.rm = T))

kable(allTracts.Summary) %>%
  kable_styling() %>%
  footnote(general_title = "\n",
           general = "Table 2.2")
```

Let's make some comparisons and speculate about the willingness to pay
and demographics in these areas 2009-2020 (see the 2000 data in the text
too)

Notice how we pipe the kable() command here

```{r}
allTracts.Summary %>%
  unite(year.TOD, year, TOD, sep = ": ", remove = T) %>%
  gather(Variable, Value, -year.TOD) %>%
  mutate(Value = round(Value, 2)) %>%
  spread(year.TOD, Value) %>%
  kable() %>%
  kable_styling() %>%
  footnote(general_title = "\n",
           general = "Table 2.3")
```

### TOD Indicator Plots

Let's create small multiple plots We use the "gather" command (look this
one up please) To go from wide to long Why do we do this?? Notice we can
"pipe" a ggplot call right into this operation!

```{r}
allTracts.Summary %>%
  gather(Variable, Value, -year, -TOD) %>%
  ggplot(aes(year, Value, fill = TOD)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~Variable, scales = "free", ncol=5) +
  scale_fill_manual(values = c("#bae4bc", "#0868ac")) +
  labs(title = "Indicator differences across time and space") +
  plotTheme() + theme(legend.position="bottom")
```

#### Examining three submarkets

```{r}
centerCity <-
  st_intersection(
    st_buffer(filter(septaStops, Line == "El"), 2640) %>% st_union(),
    st_buffer(filter(septaStops, Line == "Broad_St"), 2640) %>% st_union()) %>%
  st_sf() %>%
  mutate(Submarket = "Center City")

el <-
  st_buffer(filter(septaStops, Line == "El"), 2640) %>% st_union() %>%
  st_sf() %>%
  st_difference(centerCity) %>%
  mutate(Submarket = "El")

broad.st <-
  st_buffer(filter(septaStops, Line == "Broad_St"), 2640) %>% st_union() %>%
  st_sf() %>%
  st_difference(centerCity) %>%
  mutate(Submarket = "Broad Street")

threeMarkets <- rbind(el, broad.st, centerCity)
```

You can then bind these buffers to tracts and map them or make small
multiple plots

```{r}
allTracts.threeMarkets <-
  st_join(st_centroid(allTracts), threeMarkets) %>%
  st_drop_geometry() %>%
  left_join(allTracts) %>%
  mutate(Submarket = replace_na(Submarket, "Non-TOD")) %>%
  st_sf() 
```

### Example of using `multipleRingBuffer()` function

There is a more comprehensive tutorial on the usage of this function (and the bugs you might encounter) in the Github folder!

```{r}
allTracts <- rbind(tracts20,tracts16)
septa_MRB <- multipleRingBuffer(st_union(septaStops), 47520, 2640)

allTracts.rings <-
  st_join(st_centroid(dplyr::select(allTracts, GEOID, year)),
          septa_MRB) %>%
  st_drop_geometry() %>%
  left_join(dplyr::select(allTracts, GEOID, MedRent, year), 
            by=c("GEOID"="GEOID", "year"="year")) %>%
  st_sf() %>%
  mutate(distance = distance / 5280) #convert to miles



  ggplot() +
    geom_sf(data=septa_MRB) +
    geom_sf(data=septaStops, size=1) +
    geom_sf(data=st_union(tracts20), fill=NA, size=1.2) +
    labs(title="Half mile buffers") +
    mapTheme()

```

